---
title: "Model Assessment"
---

## Nonparametric Models

A **Plotting Position** is a non-parametric estimator of exceedance probabilities.
By using the plotting position, we can evaluate the quality of a stationary parametric model. 
To compute the plotting position, arrange the sample observations in descending order of magnitude: $x_{n:n} \geq \dots \geq x_{1:n}$.
Then, the exceedance probabilities are given by the following formula:

$$
p_{i:n} = \frac{i-a}{n+1 - 2a}, \quad i \in \{1, \dots , n\}
$$

The coefficient $a$ depends on the plotting position formula:

| Formula    | $a$     | Simplified Equation                  |
| ---------- | ------- | ------------------------------------ |
| Weibull    | $0$     | $p_{i:n} = \frac{i}{n +1}$           |
| Blom       | $0.375$ | $p_{i:n} = \frac{i-0.375}{n + 0.25}$ |
| Cunnane    | $0.4$   | $p_{i:n} = \frac{i-0.4}{n+0.2}$      |
| Gringorten | $0.44$  | $p_{i:n} = \frac{i-0.44}{n + 0.12}$  |
| Hazen      | $0.5$   | $p_{i:n} = \frac{i-0.5}{n}$          |

By default, the FFA framework uses the Weibull formula, which is unbiased.

## Accuracy Statistics 

### $R^2$ - Coefficient of Determination

To compute the $R^2$ statistic, we perform a linear regression of the annual maximum series data against the predictions of the parametric model at the plotting positions.
The $R^2$ statistic describes how well the parametric model captures variance in the data. 
Higher is better. The plot below shows the deviation of the model (red dots), from the data (black line).

![](../man/figures/plot-assessment.svg)

### RMSE - Root-Mean Squared Error

The `RMSE` statistic describes the average *squared* difference between the data and the predictions of the parametric model.
Lower is better.

### Bias

The `Bias` statistic describes the average difference between the data and the predictions of the parametric model.
A positive bias indicates that the model tends to *overestimate* the data while a negative bias indicates that the model tends to *underestimate* the data.

## Information Criterion 

The **Akaike Information Criterion** (`AIC`) and **Bayesian Information Criterion** (`BIC`) describe the quality of a model based on the error (`RMSE`) and the number of parameters (`n_theta`).
Better models have a lower `AIC`/`BIC`, which indicates that they have *less* parameters and *lower* error.

```{r, eval = FALSE}
AIC <- (n * log(RMSE)) + (2 * n_theta)
BIC <- (n * log(RMSE)) + (log(n) * n_theta)
```

The Akaike/Bayesian information criterion can also be computed using the maximum log-likelihood from [maximum likelihood estimation](parameter-estimation.md#maximum-likelihood-mle). 
These statistics are reported as `AIC_MLL` and `BIC_MLL`.

```{r, eval = FALSE}
AIC_MLL <- (n * log(MLL)) + (2 * n_theta)
BIC_MLL <- (n * log(MLL)) + (log(n) * n_theta)
```

## Uncertainty Statistics

The FFA framework uses three statistics to assess the uncertainty in flood quantile estimates:

- `AW` captures *precision* (narrower confidence intervals are better).
- `POC` captures *reliability* (higher coverage of observations is better).
- `CWI` is a composite measure balancing both precision and reliability (lower is better).

We use these metrics together to evaluate the robustness of the flood frequency analysis.

### AW – Average Width

`AW` is the average width of the interpolated confidence intervals across return periods of interest.
A smaller `AW` indicates more *precise* quantile estimates.
To compute `AW`, we use log-linear interpolation to estimate the confidence intervals of the exceedance probabilities from the confidence intervals computed during [uncertainty quantification](uncertainty-quantification.md).

### POC – Percent of Coverage

`POC` is the percentage of data points that fall within their corresponding confidence intervals.
A higher `POC` indicates greater *reliability* of the confidence intervals.

### CWI – Confidence Width Indicator

`CWI` is a composite metric that penalizes wide and/or poorly calibrated confidence intervals.

- A lower `CWI` is better.
- Wide intervals and low coverage increase the penalty.
- Ideal confidence intervals are both narrow and well-calibrated, resulting in a low `CWI`.

The `CWI` is computed using the following formula, where `alpha` is the significance level.

```{r, eval = FALSE}
CWI <- AW * exp((1 - alpha) - POC / 100)^2
```

## Handling Nonstationarity

When working with nonstationary models, adjustments must be made to both the empirical quantiles (derived from the plotting positions) and the theoretical quantiles.
In the FFA framework, we perform nonstationary model assessment by standardizing the quantiles based on the selected distribution family. 

Standardizing the theoretical quantiles:

1. Get nonexceedance probabilities $p_{(1)} \leq \dots \leq p_{(n)}$ using any plotting position formula.
2. Compute the normalized theoretical quantiles $z^{T}_{i} = \Phi^{-1}(p_{(i)})$.

Standardizing the empirical quantiles: 

1. Transform each observation $x_{i}$ into a model-based probability $u_{i}$:

    $$
    u_i = F(x_i \,;\, \mu(t_i), \sigma(t_i), \kappa)
    $$ 

2. Normalize the model-based probabilities: 

    $$
    z_{i} = \Phi^{-1}(u_{i}) 
    $$

3. Sort the normalized empirical quantiles: $z^{E}_{(1)} \leq \dots \leq z^{E}_{(n)}$.

Then, we can plot the $(z_{i}^{T}, z_{i}^{E})$ pairs in a *normalized Q-Q plot*.
Alternatively, we can use a *detrended Q-Q plot* (also known as a *worm plot*), which plots the theoretical quantiles $z_{i}^{T}$ against the differences $\Delta_{i} = z_{i}^{E} - z_{i}^{T}$.
If the model accurately captures the true distribution of the data, then:

$$
\text{SE}(\Delta_{i})= \frac{1}{f(z_{i}^{T})} \sqrt{\frac{p_{(i)}(1 - p_{(i)})}{n}}
$$ 

We can use this formula to plot the 95% confidence interval at $\pm 1.96 \cdot \text{SE}(\Delta _{i})$.
